---
title: The Odds Cav Defies
author: null
date: '2021-07-19'
slug: cav-defies-odds
categories: []
tags: []
description: ''
keywords: []
editor_options: 
  chunk_output_type: console
---

On Sunday June 28, Mark Cavendish won a Tour de France stage for the first time since 2016. A week later, he did the unthinkable and matched Merckx's record of 34 wins. As a 36 year old who had battled the Epstein-Barr virus, as well as other hurdles, few would have bet on Cav participating in the 2021 Tour de France. Fewer would have gambled on his chances to win a sprint stage against a younger generation of riders. 

What Cavendish did this summer was unbelievable. It's worth investigating what makes his path to a record-equalling 34 stages unique.




```{r data, echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(ggpattern)
library(ggtext)
library(magick)


desc_stage <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-07/tdf_stages.csv") %>%
  mutate(Type = case_when(
    str_detect(Type,"(F|f)lat|Plain")== TRUE ~ "Flat Stage",
    str_detect(Type,"Hilly|(M|m)ountain")== TRUE ~ "Mountain Stage",
    str_detect(Type, "trial")== TRUE ~ "Time Trial",
    TRUE ~ as.character(Type)
                          ))

twenty_one <- cbind(Stage = c(1,2,3,4),
              Date = c("2021-06-29","2021-07-01","2021-07-06","2021-07-09"),
              Distance = rep(100,4),
              Origin = c("Redon","Tours","Albertville","Nimes"),
              Destination = c("Fougeres","Chateauroux","Valence","Carcassonne"),
              Type = rep("Flat Stage",4),
              Winner = rep("Mark Cavendish",4),
              Winner_Country = c("GBR","GBR","GBR","GBR")) %>% as_tibble() %>%
  mutate(Date = as.Date(Date),
         Distance = as.double(Distance))

desc_stage <- rbind(desc_stage,twenty_one)
```


```{r introducing to n, echo=FALSE}
link_to_img <- function(x, width = 60) {
  glue::glue("<img src='{x}' width='{width}'/>")
}

eddies <- c("merckx-circle.png","cav-circle.png","hinault-circle.png","leducq-circle.png","darrigade-circle.png")

desc_stage  %>% count(Winner,sort = T) %>% slice(1:5) %>%
  mutate(label = link_to_img(eddies)) %>%
  ggplot(aes(fct_reorder(Winner,n),n)) + geom_col(width = .3,
                                                  fill = "yellow",
                                                  color = "black",
                                                  size = .3) + coord_flip() + 
  scale_y_continuous(breaks = seq(0,35,by=5)) +
  theme(axis.text = element_text(size=12,
                                 family = "Open Sans",
                                 color = "black"),
        panel.background = element_rect(fill="white"),
        axis.ticks.y = element_blank(),
        axis.title = element_text(hjust = 1,
                                  family = "Open Sans",
                                  size = 12)) + 
  geom_richtext(aes(label = label),
                fill = NA,
                label.color = NA,
                size = 1) +
  labs(x = NULL,
       y = "# of wins") +
  expand_limits(y = c(0,40))


```

Of the top 5 riders with the most stage wins, Cavendish is the only one still competing at the professional level. The last to retire was Merckx in 1987. Hinault, Leducq, and Darrigade have long since stopped competing.

## The Sprint Specialist

One of the things that sets Cavendish apart from others is just how specialized he is. Known as 'the manx missile' for his sprinting ability, he and his teams brought in an era of flat race finishes characterized by multiple, multi-man 'lead-out trains'. In fact, all of his stages won are on flat stages. This differs greatly from Merckx and the next 3 winningest riders, who have won time trials as well as mountain stages.

```{r type spread, fig.align="center", echo=FALSE,message=FALSE}

desc_stage %>% 
  filter(str_detect(Winner,
                    "Cavendish|Hinault|Merckx|André Leducq|André Darrigade")) %>% 
  filter(!str_detect(Winner,"Antonin")) %>%
  group_by(Winner,Type) %>% summarise(wins = n()) %>% ungroup() %>%
  group_by(Winner,Type) %>% summarise(wins = sum(wins)) %>% ungroup() %>%
  group_by(Winner) %>% mutate(perc = wins/sum(wins),
                              total_wins = sum(wins)) %>% ungroup() %>%
  ggplot(aes(Winner,perc,fill=Type)) + geom_col_pattern(aes(pattern_density=Type),
                                                        pattern='circle',
                                                        pattern_fill="red",
                                                        pattern_color="red",
                                                        color = "black",
                                                        size=.3) + coord_flip() + 
  theme(axis.text = element_text(size=12,
                                 family = "Open Sans",
                                 color = "black"),
        panel.background = element_rect(fill="white"),
        axis.ticks = element_blank(),
        axis.title = element_text(hjust = 1,
                                  family = "Open Sans",
                                  size = 12),
        legend.position = "bottom",
        axis.text.x = element_blank(),
        legend.text = element_text(family="Open Sans"),
        legend.title = element_text(family = "Open Sans")) +
  scale_fill_manual(values = c("#4ba82e","white","yellow"))+
labs(y = "",
     x = "") + 
  scale_pattern_density_manual(values = c(0,0.5,0))

```

## The Longest Road

Part of the revelation of Cav's resurgence is the fact that he is 36 year old, over a decade removed from his first tour appearance. Cav's specialization in flat, bunch sprint stages has meant he's only has a few opportunities to win each year, making his record-setting journey a long one. His comeback in 2021 also bucks the trend of the other top winnners, who've never had more than a single gap year between their first and last tour victories.

```{r length of stages,fig.align="center", echo=FALSE,message=FALSE,fig.height=7}
library(lubridate)

wins_alter <- desc_stage %>% 
  filter(str_detect(Winner,
                    "Cavendish|Hinault|Merckx|André Leducq|André Darrigade"),
         !str_detect(Winner,"Antonin")) %>% 
  select(Winner,Date,Type) %>%
  mutate(year = year(Date)) %>%
  group_by(Winner) %>% mutate(distance = year-min(year)) %>% ungroup() %>% 
  group_by(Winner,year,distance,Type) %>% summarise(num_wins = n()) 


desc_stage %>% 
  filter(str_detect(Winner,
                    "Cavendish|Hinault|Merckx|André Leducq|André Darrigade"),
         !str_detect(Winner,"Antonin")) %>%
  group_by(Winner) %>% summarise(end = max(Date),
                                 start = min(Date)) %>% ungroup() %>%
  mutate(gap = year(end)-year(start),
         beg = 0)%>%
  ggplot(aes(y= Winner,xend=gap,x=beg,yend=Winner))  +
  geom_jitter(width = .25,
              height=.23,
              inherit.aes = F,data = wins_alter,aes(distance,Winner,size=num_wins,fill=Type),
              pch = 21,
              color = "black") + 
  scale_size(range = c(3, 15)) + 
  scale_x_continuous(breaks = seq(0,13,by=1)) + 
  theme(axis.text = element_text(size=12,
                                 family = "Open Sans",
                                 color = "black"),
        panel.background = element_rect(fill="white"),
        axis.ticks = element_blank(),
        axis.title = element_text(hjust = 1,
                                  family = "Open Sans",
                                  size = 12),
        legend.position = "bottom",
        legend.text = element_text(family="Open Sans"),
        legend.title = element_text(family = "Open Sans"),
        panel.grid.major.x = element_line(size = .2,
                                          linetype = "dashed",
                                          color = "black")) +
  scale_fill_manual(values = c("#4ba82e","red","yellow")) +
  labs(x = "# years since first tdf win",
       y = NULL,
       fill = NULL)  +
  guides(size = "none")

```

Merckx and Cavendish are clearly different in that the first record holder won his stages in a rather short amount of time. This year's tour has Mark taking the longest journey of stage winning, at 13 years. However, the two are the same in that if they won once, they were going to win again. 

Going forward, it's worth asking what will be required for a new rider to beat or match Cav and Merckx's record. As some of us have seen over the course of Cavendish's career, there was a large amount of team commitment to get Mark to the line first. Racers like Hinault and Merckx have built their legacies from stage wins that required them to make their own victories. As both of them are also some of the winningest men of the overall Tour de France, it may take a hegemonic grand tour rider to follow their in footsteps.

One thing is undeniable: Mark Cavendish has achieved an unexpected match to Merkckx's record by paving a way much different than those who came before him.
